BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2014-07-30 Paper/BOAST.tex
Improved Paper

>>>>

Usually computing kernels are hotspots of an HPC application, and are most of
the time based on a loop nest. A lot of effort is dedicated to their tuning and
the obtained result is often quite different than the original procedure.
Several transformations can be applied to such kernels. And those optimization
are often applied manually as compiler may fail to recognise the opportunity.

There are many different loop optimization techniques~\cite{wolf1991loop}. We
can cite loop skewing~\cite{wolfe1986loops} (derives nested loops wavefronts)
or loop interchange~\cite{allen1984automatic} (loop variables change places).

The importance of  correct loop imbrication on BLAS~\cite{lawson1979basic}
operations is studied in~\cite{soliman2009performance}, and shows performance
increase of a factor up to 5 when using correct loop imbrication. The importance
of code transformation is stressed in~\cite{ye2011porting}, where a selection of GPU
kernels are ported to CPU and optimized.

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite FORTRAN and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions must be composed in complex
Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance commonly used operator have to be available
and behave as expected. It must also be possible to differentiate an action on
an abstraction in the context of an expression and in the context of the
management of this expression. This is why an embedded domain specific language
approach was selected~\cite{hudak1996building}. This allows for the coexistence
of two languages: the host language and the DSL. In our case the DSL allows the
description of the kernel while the host language provides the meta programming
of the kernel. Operator overloading of the host language will provide the
familiar syntax programmers are accustomed to.

It was also important to have our construct like for loops to have a syntax
approaching those commonly found in programming languages. For this we needed a
language which could seamlessly pass a block of code to a function.
Ruby~\cite{matsumoto2002ruby} is one such language. It has deep introspection
capabilities as well.

%Considering source-to-source transformations that we investigate with the BOAST
%tool in this paper, we can compare ourselves to POET~\cite{POET2012}  and
%ABCLibScript~\cite{ABCLib2006}. However, both tools apply transformations
%to the syntaxic tree of a program while BOAST intervenes well before in the
%transformation process. Indeed, one may first use BOAST on order to generate a
%program (a specific syntaxic tree) and then benefit from the transformation
%facilities offered by one of the above tools.
